### 2016-05-24 thesis_paper/source.tex
changes to thesis paper

\linespread{1.5}

\titleformat{\chapter}
{\normalfont\LARGE\bfseries}{\thechapter.}{1em}{}

@@

@@

\begin{abstract}
\justify
\hspace{3em} Modern simulation systems generate big amount of data, which consequently has to be analyzed in a timely fashion. Traditional database management systems follow principle of pulling the needed data, processing it, and then returning the results. This approach is then optimized by means of caching, storing in different structures, or doing some sacrifices on precision of the results to make it faster. When it comes to the point of doing various queries that require analysis  of the whole data,  this design has the following disadvantages: considerable overhead on traditional disk random I/O framework while reading from the simulation output files and low throughput of the data that consequently results in long latency, and, if there was any indexing to optimize selections, overhead of storing those becomes too big, too. Besides it, indexing will also cause delay during write operations and since most of the queries work with the entire data sets, indexing loses its point.

@@

\begin{figure}
 \centerline{ \includegraphics[width=0.5\columnwidth]{images/sample_snapshot_of_simulation.eps} }
 \caption{ Snapshots of two MS systems: a collagen fiber structure with 890,000 atoms (top) and a dipalmitoylphosphatidylcholine (DPPC) bi-layer lipid system with 402,400 atoms (bottom) \cite{mainArticle}}
 \label{fg:collagen_dipal}
\end{figure}

@@

\begin{figure}
 \centerline{ \includegraphics[width=0.5\columnwidth]{images/speedup838K-pomalo-za-4.eps} }
 \caption{ Speed up over different levels of atom selection for 838,400 atoms \cite{mainPaper}}
 \label{fg:sample_estimation_old_paper}
\end{figure}

@@

@@

We are potentially dealing with data that could easily get up to terabytes and petabytes in sizes, and it will be fair to apply term of big data. Nowadays there are bunch of frameworks that try to solve this problem. Some of these frameworks are Hadoop \cite{hadoop}, Hadoop with GPU extension \cite{hadoopgpu}, Spark \cite{spark}, Flink \cite{flink} (this is more of a big data stream processing framework) and other ones that are possibly just custom modifications and distributions of these. The mentioned tools help dealing with big data giving an ability to process them in batches on clusters using underlying HDFS (Highly Distributed File System), and normally go along with other utilities that help to apply MapReduce concept. In fact, the frameworks listed above could even be use for our problem on top of our system, because, as we discuss in section \ref{sc:conclusion}, it could be extended to a distributed system for analyzing independent frames in parallel.

@@

Since dealing with data is very delicate in this kind of engines, there are couple of features and requirements that this architecture is applicable to, in order to avoid inappropriate understanding or application of this approach in systems that were covered in the lectures\cite{DataStreamManagement}. The data is volatile and not persistent anymore because of obvious sizes of data.  Since we are pulling the data as little times as possible taking into consideration that MS data is only appended and not modified, the access is sequential rather than random. There are concepts of framing and windowing, and memory is limited in streaming engine, while in DBMS secondary storage is considered to be unlimited. Obviously, because of all of these features or requirements to data, the streaming engine frameworks have some limitations like going back into the history of data, since it is volatile. In fact, this has also been tried by Borealis Streaming Engine \cite{borealis}. On top of the fact that it is a streaming engine, as it was built based on Medusa\cite{medusa} and Aurora\cite{aurora}, it proposes distributed processing at this point having some features to handle errors and look back in the history. In our case, it is primarily a single data source, since it is based on the simulation, though in the future the simulation software systems might introduce distributed computing, too. It is also important to mention such projects as PeriScope\cite{PeriScope}, SciDB\cite{SciDB}, and other\cite{SDSS_SIGMOD02,QBISM_ICDE94}, since they target scientific data, which is a lot more different, than usual industry customer oriented data.

@@

\chapter{Network of Queries}\label{sc:querynetwork}
Having generated big amount of data in large files, now scientists need to get useful information out of it by means of analysis utilities. Often times, the queries that need to be run over the data look like selection and some kind of accumulation, if it is not more complicated. To be more exact, in this section we will try to summarize common set of queries we will be improving.

In previous work, there has already been developed a network of queries widely used by scientists for MS systems. In Table~\ref{tb:queries} you may observe the common set of queries that has been developed. Basically, these are the queries that need to be improved with our new approach of parallelism. Just to mention, $n$, $r_i$, $m_i$, $c_i$ and $q_i$ denote number of particles, coordinates (vector form), mass, charge, and number of electrons of a particle $i$.

\begin{table}[h]
\resizebox{\columnwidth}{!}{\begin{minipage}{\columnwidth}
\renewcommand*{\arraystretch}{1.5}
\tabcolsep=0.12cm
\begin{tabular}{|c| c|}
\hline %inserts horizontal line
Function Name & Equation/Description \\[0.5ex] \hline
Moment of Inertia & $\begin{array} {lcl} I & = & \sum\limits_{i=1}^n m_ir_i \end{array}$ \\[0.5ex]
\hline
Moment of Inertia on z axis & $\begin{array} {lcl} I_z & = & \sum\limits_{i=1}^n m_ir_{zi} \end{array}$ \\[0.5ex]
\hline
Sum of masses & $\begin{array} {lcl} M & = & \sum\limits_{i=1}^n m_i \end{array}$ \\[0.5ex]
\hline
Center of mass & $\begin{array} {lcl} CoM & = & \frac{I}{M} \end{array}$ \\[0.5ex]
\hline
Radius of Gyration & $\begin{array} {lcl} RG & = & \sqrt{\frac{I_z}{M}} \end{array}$ \\[0.5ex]
\hline
Dipole Moment& $\begin{array} {lcl} D & = & \sum\limits_{i=1}^n q_ir_i \end{array}$ \\[0.5ex]
\hline
Dipole Histogram & $\begin{array} {lcl} D_z & = & \sum\limits_{i=1}^n \frac{D}{z} \end{array}$ \\[0.5ex]
\hline
Electron Density & $\begin{array} {lcl} ED & = & \frac{\sum\limits_{i=1}^n (e_i-q_i)}{dz \cdot x \cdot y} \end{array}$ \\[0.5ex]
\hline
Heat Capacity & $\begin{array} {lcl} HC & = & \frac{3000 \cdot \sqrt{T} \cdot boltz}{2 \cdot \sqrt{T}-n \cdot df \cdot VarT} \end{array}$ \\[0.5ex]
\hline
Isothermal Compressibility & $\begin{array} {lcl} I & = & \frac{VarV}{V_{avg} \cdot boltz \cdot T \cdot PresFac} \end{array}$ \\[0.5ex]
\hline
Mean Square Displacement& $\begin{array} {lcl} msd & = & \langle (r_{t+\Delta_t}-r_t)^2 \rangle \end{array}$ \\[0.5ex]
\hline
Diffusion Constant & $\begin{array} {lcl} D_t & = & \frac{6 \cdot msd(t)}{t} \end{array}$ \\[0.5ex]
\hline Velocity Autocorrelation & $\begin{array} {lcl} V_{acor} & =&
\langle (V_{t+\Delta_t}\cdot V_t) \rangle \end{array}$ \\[0.5ex]
\hline
Force Autocorrelation & $\begin{array} {lcl} F_{acor} & = & \langle (F_{t+\Delta_t} \cdot F_t) \rangle \end{array}$ \\[0.5ex]
\hline
Density Function & Histogram of atom counts \\[0.5ex]
\hline
SDH & Histogram of all distances \\[0.5ex]
\hline
RDF & $\begin{array} {lcl} rdf(r) & = & \frac{SDH(r)}{4 \cdot \pi \cdot r^2 \cdot \sigma_r \cdot \rho} \end{array}$ \\[0.5ex]
\hline
\end{tabular}
\caption[Table caption text]{Popular analytical queries in MS\cite{mainPaper} }
\label{tb:queries}
\end{minipage} }
\end{table}

In general, with the given queries we have divided them into two categories: one body functions and multiple body functions. The first ones apparently have linear complexity of $O(n)$, while the other ones have bigger complexity.

@@

@@

The might be some dependencies and preliminary computation possibilities. For example, some queries need total mass, and we could compute it before we push it further, but this can be easily done by user based on the complexity and dependency. It is explained in more details in further sections.

@@

In order to experiment with different input cases, the input files have been generated with a tool that we have built in Python language. Thus, we are able to generate mock data files with any number of frames and atoms, and the values are random and can be adjusted with needs of the experiment (for example, for cases when we want to analyze bucket widths in a limited space, the coordinates are generated in range of $0 - 1$, etc.). As it has already been mentioned earlier, we don't expect the data to be distributed with a certain behavior, the importance of having this kind of tool to generate mock data is having a variety of random data sets that could reveal any issues related to values provided to the program, on top of that, having the actual file written and then being read has been a part of the experiments to have more realistic use case. In the future, the same script for generating the mock data could be used for regression and performance tests in a distributed implementation of this design as a single source.

@@

MS data roughly can be represented as a list of atoms with physical properties aggregated as frames. Thus, we consider a single frame as a chunk which will be loaded into the memory\footnote{Memory in this design might be interpreted differently, since we have a separate device memory on GPUs, but it will be explained better later} and pushed through our network of query. In this case, it is very convenient, because there are almost no dependencies between frames, rather than list of atoms in the same frame, which potentially again could create additional disk I/O overhead.

@@

@@

\emph{\textbf{GPU execution}}. From hardware standpoint, it is important to understand that every GPU device consists of SM (streaming multiprocessors), and each SM has thousands of registers, which can be partitioned among execution threads. Having this in mind, since this is a pure parallel processing power, the goal is to increase hardware throughput and keep highest utilization to exploit every register given. CUDA is an amazing tool that tries to help developers map this hardware capabilities with the actual application layer. In CUDA we have grids, blocks, and threads, where grid is 3 dimensional unit where each cell is a block, and the same relationship applies between blocks and threads\footnote{This explanation doesn't include memory setup in GPU devices, but it is covered in further sections}. Originally, this mapping was perfect for solving image processing problems or other graphics computation, since dimensions and such division (e.g. by pixels to threads) is very inherent. As a language feature, in CUDA we have to define such a function called 'kernel' function. Essentially, it is the function where we are supposed to map every execution thread with data and perform the actual execution, this approach is widely known as SIMD\footnote{SIMD - Single Instruction Multiple Data}. As a convenience we have separated one-body functions with multiple-body functions into separate kernels.

@@

\begin{figure}
 \centerline{ \includegraphics[width=0.65\columnwidth]{images/reduction} }
 \caption{Array reduction example\cite{gpureduction}}
 \label{fg:reduction}
\end{figure}

@@

The framework has been developed in a programming language CUDA, and the sequential version for benchmark\cite{mainPaper} has been originally developed C++. Experiments were run on operating systems which didn't have any additional workload besides usual installation, and the operating systems particularly were Amazon Linux, which doesn't have a lot of modifications except for some additional features for management with Amazon EC-2 Console. We have generated mock data sets with different variations of atoms, frames and ran them with different workloads comparing sequential version of Push-based system with improved version written in CUDA.

\emph{\textbf{Hardware stack.}}
There are primarily several parts that are really important in this hardware stack. Since this is primarily big data processing, we care about processing power and memory. RAM on the node that we were running has 16GB. The server allocated from Amazon has the following CPU parameters:

\begin{center}
  \begin{tabular}{ l  l }

Architecture & x86\_64 \\  
CPU op-mode(s) & 32-bit, 64-bit \\  
Byte Order & Little Endian \\  
CPU(s) & 8 \\  
On-line CPU(s) list & 0-7 \\  
Thread(s) per core & 2 \\  
Core(s) per socket & 4 \\  
CPU family & 6 \\  
Model & 45 \\  
Model name & Intel Xeon E5-2670 2.60GHz \\  
Stepping & 7 \\  
CPU MHz & 2593.814 \\  
L1d cache & 32K \\  
L1i cache & 32K \\  
L2 cache & 256K \\  
L3 cache & 20480K \\  
NUMA node0 CPU(s) & 0-7 \\

  \end{tabular}
\end{center}

@@

\begin{center}
  \begin{tabular}{ l l }
Name & Grid 34C K520 \\
Memory & 4095Mib \\ 
Total Nvidia CUDA Cores & 1536 \\ 
Memory Size & 8 GB GDDR5 (4 GB/GPU) \\
Total GPUs & 2 GK104 GPUs \\
  \end{tabular}
\end{center}

@@

>>>>

\linespread{2}

\titleformat{\chapter}[display]
        {\normalfont\Large\bfseries\centering}{\chaptertitlename\ \thechapter}{0pt}{\Large\uppercase}

\renewcommand{\contentsname}{CONTENTS}
\renewcommand{\listfigurename}{LIST OF FIGURES}
\renewcommand{\listtablename}{LIST OF TABLES}
\renewcommand{\abstractname}{ABSTRACT}

% Make chapter titles uppercase is a pain.
% The internal chapter macros must be redefined.
\makeatletter
\renewcommand\chapter{\if@openright\cleardoublepage\else\clearpage\fi
                    \thispagestyle{plain}%
                    \global\@topnum\z@
                    \@afterindentfalse
                    \secdef\@chapter\@schapter}
\def\@chapter[#1]#2{\ifnum \c@secnumdepth >\m@ne
                       \if@mainmatter
                         \refstepcounter{chapter}%
                         \typeout{\@chapapp\space\thechapter.}%
                         \addcontentsline{toc}{chapter}%
                         {\protect\numberline{\thechapter}\texorpdfstring{\MakeUppercase{#1}}{#1}}%
                       \else
                       \addcontentsline{toc}{chapter}{\texorpdfstring{\MakeUppercase{#1}}{#1}}%
                       \fi
                    \else
                    \addcontentsline{toc}{chapter}{\texorpdfstring{\MakeUppercase{#1}}{#1}}%
                    \fi
                    \chaptermark{#1}%
                    \addtocontents{lof}{\protect\addvspace{10\p@}}%
                    \addtocontents{lot}{\protect\addvspace{10\p@}}%
                    \if@twocolumn
                      \@topnewpage[\@makechapterhead{#2}]%
                    \else
                      \@makechapterhead{#2}%
                      \@afterheading
                    \fi}
\makeatother

@@

\linespread{1}

@@

\addcontentsline{toc}{chapter}{\listfigurename}
\listoffigures

\addcontentsline{toc}{chapter}{\listtablename} 
\listoftables

\linespread{2}
\clearpage

\addcontentsline{toc}{chapter}{\abstractname} 
\begin{abstract}
\justify
\hspace{3em} Modern simulation systems generate big amount of data, which consequently has to be analyzed in a timely fashion. Traditional database management systems follow principle of pulling the needed data, processing it, and then returning the results. This approach is then optimized by means of caching, storing in different structures, or doing some sacrifices on precision of the results to make it faster. When it comes to the point of doing various queries that require analysis  of the whole data,  this design has the following disadvantages: considerable overhead on traditional disk random I/O framework while reading from the simulation output files and low throughput of the data that consequently results in long latency, and, if there was any indexing to optimize selections, overhead of storing those becomes too big, too. Besides it, indexing will also cause delay during write operations and since most of the queries work with the entire data sets, indexing loses its point.

@@

\begin{figure}
 \centerline{ \includegraphics[width=0.5\columnwidth]{images/sample_snapshot_of_simulation.eps} }
 \caption{ Snapshots of two MS systems with 890,000 atoms with 402,400 atoms}
 \label{fg:collagen_dipal}
\end{figure}

@@

\begin{figure}
 \centerline{ \includegraphics[width=0.5\columnwidth]{images/speedup838K-pomalo-za-4.eps} }
 \caption{ Speed up over different levels of atom selection for 838,400 atoms}
 \label{fg:sample_estimation_old_paper}
\end{figure}

@@

As a plugin for optimizing computation of scientific queries against large data sets being produced, Push-based System for Molecular Simulation Data Analysis with GPU improvement should be able to handle data streams as well, in order to get real time results and saving up on time, instead of pulling everything from files later on.

@@

We are potentially dealing with data that could easily get up to terabytes and petabytes in sizes, and it will be fair to apply term of big data. Traditional software tools are not able to process Big Data apparently because of huge data sets, thus, special frameworks were developed to handle them. Nowadays there are bunch of frameworks that try to solve this problem. Some of these frameworks are Hadoop \cite{hadoop}, Hadoop with GPU extension \cite{hadoopgpu}, Spark \cite{spark}, Flink \cite{flink} (this is more of a big data stream processing framework) and other ones that are possibly just custom modifications and distributions of these. The mentioned tools help dealing with big data giving an ability to process them in batches on clusters using underlying HDFS (Highly Distributed File System), and normally go along with other utilities that help to apply MapReduce concept. The relationship between our system and this concept along with its implementations is in data sets sizes. Even though we are not using or introducing any non-traditional file systems, in certain scale and problems it could be required, too. In fact, the frameworks listed above could even be used for our problem on top of our system, because, as we discuss in section \ref{sc:conclusion}, it could be extended to a distributed system for analyzing independent frames in parallel.

\section{GPU and Big Data Processing}
GPU is becoming more popular being used by not only graphics processing, but machine learning, deep learning, search algorithms, sorting, and so on. Having an ability to optimize in tremendous speedups it is being involved in video classifications, natural language processing. Particularly about distributed processing and GPU applications along with it, there has been made some overview and research on what is the current state for common solutions in industry. "There were a lot of custom GPU-Hadoop frameworks after the launch of Mars project. These include C-MR, SteamMR, GPMR, Grex, Panda, Shredder and many others. However, most of these frameworks are no longer supported and were developed for some particular scientific projects. Therefore, a Monte Carlo simulation framework can be hardly applied for a bioinformatics project (say) based on the other algorithms. Moreover, processor technologies are evolving very rapidly. Many new revolutionary architectures have been developed for Sony PlayStation 4, Mali GPU by ARM, Adapteva Multicore Microprocessor etc. Both Mali GPU and Adapteva will be compatible with OpenCL. Xeon Phi co-processor works with OpenCL too, which was launched by INTEL. It has an x86-like architecture and is a 60 core co-processor that supports the PCI Express standard. It consumes just 300watts of energy and provides by giving a performance of 1 TFLOPS in double precision. Tianhe-2 which is the most powerful supercomputer till date implements this co-processor. Though, it is very difficult determine which Processor Computing framework architecture will more robust and high
performance." \cite{hadoopgpu}

@@

Since dealing with data is very delicate in this kind of engines, there are couple of features and requirements that this architecture is applicable to, in order to avoid inappropriate understanding or application of this approach in systems that were covered in the lectures\cite{DataStreamManagement}. The data is volatile and not persistent anymore because of obvious sizes of data.  Since we are pulling the data as little times as possible taking into consideration that MS data is only appended and not modified, the access is sequential rather than random. There are concepts of framing and windowing, and memory is limited in streaming engine, while in DBMS secondary storage is considered to be unlimited. Obviously, because of all of these features or requirements to data, the streaming engine frameworks have some limitations like going back into the history of data, since it is volatile. In fact, this has also been tried by Borealis Streaming Engine \cite{borealis}. On top of the fact that it is a streaming engine, as it was built based on Medusa\cite{medusa} and Aurora\cite{aurora}, it proposes distributed processing at this point having some features to handle errors and look back in the history. In our case, it is primarily a single data source, since it is based on the simulation, though in the future the simulation software systems might introduce distributed computing, too.

@@

\chapter{Benchmark scientific queries}\label{sc:querynetwork}
Having generated big amount of data in large files, now scientists need to get useful information out of it by means of analysis utilities. Often times, the queries that need to be run over the data look like selection and some kind of accumulation, if it is not more complicated. To be more exact, in this section we will try to summarize common set of queries that were already introduced in the previous work of Push-Based System for Molecular Simulation Data Analysis \cite{mainPaper}.

In previous work, there has already been developed a network of queries widely used by scientists for MS systems. In Table~\ref{tb:queries} you may observe the common set of queries that has been developed. Basically, these are the queries that need to be improved with our new approach of parallelism. Respectively, $n$, $r_i$, $m_i$, $c_i$ and $q_i$ denote number of particles, coordinates (vector form), mass, charge, and number of electrons of a particle $i$.

\begin{table}[h]
\centering
\resizebox{\columnwidth}{!}
{
\begin{minipage}{\columnwidth}
\centering
\renewcommand*{\arraystretch}{1.5}
\tabcolsep=0.12cm
\begin{tabular}{|c| c|}
\hline %inserts horizontal line
Function Name & Equation/Description \\[0.5ex] 
\hline
Moment of Inertia & $\begin{array} {lcl} I & = & \sum\limits_{i=1}^n m_ir_i \end{array}$ \\[0.5ex]
\hline
Moment of Inertia on z axis & $\begin{array} {lcl} I_z & = & \sum\limits_{i=1}^n m_ir_{zi} \end{array}$ \\[0.5ex]
\hline
Sum of masses & $\begin{array} {lcl} M & = & \sum\limits_{i=1}^n m_i \end{array}$ \\[0.5ex]
\hline
Center of mass & $\begin{array} {lcl} CoM & = & \frac{I}{M} \end{array}$ \\[0.5ex]
\hline
Radius of Gyration & $\begin{array} {lcl} RG & = & \sqrt{\frac{I_z}{M}} \end{array}$ \\[0.5ex]
\hline
Dipole Moment& $\begin{array} {lcl} D & = & \sum\limits_{i=1}^n q_ir_i \end{array}$ \\[0.5ex]
\hline
Dipole Histogram & $\begin{array} {lcl} D_z & = & \sum\limits_{i=1}^n \frac{D}{z} \end{array}$ \\[0.5ex]
\hline
Electron Density & $\begin{array} {lcl} ED & = & \frac{\sum\limits_{i=1}^n (e_i-q_i)}{dz \cdot x \cdot y} \end{array}$ \\[0.5ex]
\hline
Heat Capacity & $\begin{array} {lcl} HC & = & \frac{3000 \cdot \sqrt{T} \cdot boltz}{2 \cdot \sqrt{T}-n \cdot df \cdot VarT} \end{array}$ \\[0.5ex]
\hline
Isothermal Compressibility & $\begin{array} {lcl} I & = & \frac{VarV}{V_{avg} \cdot boltz \cdot T \cdot PresFac} \end{array}$ \\[0.5ex]
\hline
Mean Square Displacement& $\begin{array} {lcl} msd & = & \langle (r_{t+\Delta_t}-r_t)^2 \rangle \end{array}$ \\[0.5ex]
\hline
Diffusion Constant & $\begin{array} {lcl} D_t & = & \frac{6 \cdot msd(t)}{t} \end{array}$ \\[0.5ex]
\hline Velocity Autocorrelation & $\begin{array} {lcl} V_{acor} & =&
\langle (V_{t+\Delta_t}\cdot V_t) \rangle \end{array}$ \\[0.5ex]
\hline
Force Autocorrelation & $\begin{array} {lcl} F_{acor} & = & \langle (F_{t+\Delta_t} \cdot F_t) \rangle \end{array}$ \\[0.5ex]
\hline
Density Function & Histogram of atom counts \\[0.5ex]
\hline
SDH & Histogram of all distances \\[0.5ex]
\hline
RDF & $\begin{array} {lcl} rdf(r) & = & \frac{SDH(r)}{4 \cdot \pi \cdot r^2 \cdot \sigma_r \cdot \rho} \end{array}$ \\[0.5ex]
\hline
\end{tabular}
\caption[Popular analytical queries in MS]{Popular analytical queries in MS }
\label{tb:queries}

\end{minipage}

}
\end{table}

The queries that we consider are related to computations limited within one frame of a molecular simulation. In general, with the given queries we have divided them into two categories: one body functions and multiple body functions. The first ones apparently have linear complexity of $O(n)$, while the other ones have bigger complexity.

@@

As you can see in the given table, most of them are basically based on primitive summation or other basic mathematical operations, in our experiments in further sections we will be running result on single precision values or to be more explicit we use float type in C programming language.

@@

For computation of histogram it is not enough to visit every value for each atom once, since we have to do pairwise computations, thus it increases complexity of the problem not only computation-wise, but memory-wise as well, which is a bigger concern in our case, since the overhead will be huge and non-linearly relative to the size of data. For this purpose we have come up with a way to optimize the way we use references to memory for each atom, so that we minimize overhead of pulling data all the way from DRAM (GPU).

There might be some dependencies and preliminary computation possibilities. For example, some queries need total mass, and we could compute it before we push it further, but this can be easily done by user based on the complexity and dependency. It is explained in more details in further sections.

@@

In order to experiment with different input cases, the input files have been generated with a tool that we have built in Python language. Thus, we are able to generate mock data files with any number of frames and atoms, and the values are random and can be adjusted with needs of the experiment (for example, for cases when we want to analyze bucket widths in a limited space, the coordinates are generated in range of $0 - 1$, etc.). As it has already been mentioned earlier, we don't expect the data to be distributed with a certain behavior, the importance of having this kind of tool to generate mock data is having a variety of random data sets that could reveal any issues related to values provided to the program, on top of that, having the actual file written and then being read has been a part of the experiments to have more realistic use case. In the future, the same script for generating the mock data could be used for regression and performance tests in a distributed implementation of this design as a single source. Potentially input can be technically anything including simple IO sockets that could simply the way the system communicates with other components (f.e. over network) which would also contribute to expanding it to distributed cluster processing (though we believe that it would require additional components such as scheduler or job queues, etc.). To be more exact, the input format of the data and the data represents nothing but a simple array of tuples where each element represents certain descriptor of the given atom in the current frame. We believe that this simple approach makes it a good generalization of solution which would be user friendly for any scientist that wants to run any custom analysis on his data sets.

@@

MS data roughly can be represented as a list of atoms with physical properties aggregated as frames. Thus, we consider a single frame as a chunk which will be loaded into the memory\footnote{Memory in this design might be interpreted differently, since we have a separate device memory on GPUs, but it will be explained better later} and pushed through our network of queries. In this case, it is very convenient, because there are almost no dependencies between frames, rather than list of atoms in the same frame, which potentially again could create additional disk I/O overhead.

@@

To summarize, the data goes through several levels of memory hierarchy in our system. Coming as an input from either streaming tool, flat file (our case), or other type of input (f.e. socket) which is either disk memory or RAM it goes to RAM of our system process. After it is loaded to RAM, we continue further and load it to Global Memory of GPU, meanwhile while it is loading the given chunk of data to GPU Global Memory (which happens asynchronously), we as well load the next chunk of data to RAM. In this case, it helps us avoid minor overhead of transferring data between CPU to GPU. Further, we take use of different types of memories in GPU to optimize different queries which will be described in further sections.

@@

\emph{\textbf{GPU execution}}. From hardware standpoint, it is important to understand that every GPU device consists of SMs (streaming multiprocessors), and each SM has thousands of registers, which can be partitioned among execution threads. Having this in mind, since this is a pure parallel processing power, the goal is to increase hardware throughput and keep highest utilization to exploit every register given. CUDA is an amazing tool that tries to help developers map this hardware capabilities with the actual application layer. In CUDA we have grids, blocks, and threads, where grid is 3 dimensional unit where each cell is a block, and the same relationship applies between blocks and threads\footnote{This explanation doesn't include memory setup in GPU devices, but it is covered in further sections}. Abstract dimensions given by CUDA are not real entities of physical GPU memory and its organization, but it is just a very convenient way of organizing data and tasks to help GPU scheduler pick and execute the instructions. Originally, this mapping was perfect for solving image processing problems or other graphics computation, since dimensions and such division (e.g. by pixels to threads) is very inherent. As a language feature, in CUDA we have to define such a function called 'kernel' function. Essentially, it is the function where we are supposed to map every execution thread with data and perform the actual execution, this approach is widely known as SIMD\footnote{SIMD - Single Instruction Multiple Data}. As a convenience we have separated one-body functions with multiple-body functions into separate kernels.

@@

\begin{figure}
 \centerline{ \includegraphics[width=0.65\columnwidth]{images/reduction} }
 \caption{GPU array reduction algorithm sample representation}
 \label{fg:reduction}
\end{figure}

@@

The framework has been developed in a programming language CUDA, and the sequential version for benchmark\cite{mainPaper} has been originally developed in C++. Experiments were run on operating systems which didn't have any additional workload besides usual installation, and the operating systems particularly were Amazon Linux, which doesn't have a lot of modifications except for some additional features for management with Amazon EC-2 Console. We have generated mock data sets with different variations of atoms, frames and ran them with different workloads comparing sequential version of Push-based system with improved version written in CUDA.

\begin{table}[h!]
	\renewcommand*{\arraystretch}{1.5}
	\tabcolsep=0.12cm
  \centering
  \label{tab:cpu_hardware}
  \begin{tabular}{| l | r |}
   \hline
Architecture & x86\_64 \\  \hline
CPU op-mode(s) & 32-bit, 64-bit \\   \hline
Byte Order & Little Endian \\   \hline
CPU(s) & 8 \\   \hline
On-line CPU(s) list & 0-7 \\   \hline
Thread(s) per core & 2 \\   \hline
Core(s) per socket & 4 \\   \hline
CPU family & 6 \\   \hline
Model & 45 \\   \hline
Model name & Intel Xeon E5-2670 2.60GHz \\   \hline
Stepping & 7 \\   \hline
CPU MHz & 2593.814 \\  \hline  
L1d cache & 32K \\   \hline
L1i cache & 32K \\   \hline
L2 cache & 256K \\   \hline
L3 cache & 20480K \\   \hline
NUMA node0 CPU(s) & 0-7 \\  \hline
  \end{tabular}
    \caption{Hardware stack for CPU version of Push-based system benchmark}
\end{table}
\emph{\textbf{Hardware stack.}}
There are primarily several parts that are really important in this hardware stack. Since this is primarily big data processing, we care about processing power and memory. RAM on the node that we were running has 16GB. The server allocated from Amazon has the following CPU parameters:

@@

@@

\begin{table}[h!]
	\renewcommand*{\arraystretch}{1.5}
	\tabcolsep=0.12cm
  \centering
  \label{tab:gpu_hardware}
  \begin{tabular}{| l | r |}
   \hline
	Name & Grid 34C K520 \\ \hline
	Memory & 4095Mib \\ \hline
	Total Nvidia CUDA Cores & 1536 \\ \hline
	Memory Size & 8 GB GDDR5 (4 GB/GPU) \\ \hline
	Total GPUs & 2 GK104 GPUs \\ \hline
  \end{tabular}
    \caption{Hardware stack for GPU version of Push-based system}
\end{table}
